Evolution of pre-copulatory and post-copulatory inbreeding behaviour

### 2016-09-22 ms_pre_post.Rmd
Change title: First of many changes following internal review
Shorten introduction, four paragraphs, third might still need work

---
title: 'Evolution of pre-copulatory and post-copulatory inbreeding behaviour: a computational
  approach'
author: "A. Bradley Duthie et al."
date: "August 24, 2016"
output:
  pdf_document:
    fig_caption: yes
  html_document: default
  word_document:
    fig_caption: yes
    pandoc_args:
    - --csl
    - evolution.csl
header-includes:
- \usepackage{natbib}
- \usepackage{lineno}
- \linenumbers
- \bibliographystyle{amnatnat}
linestretch: 1
csl: evolution.csl
bibliography: ms_pre_post.bib
---

@@

Reproductive decisions expressed among interacting individuals directly affect individual fitnesses and the genotypes and viabilities of resulting offspring, and thereby profoundly affect evolutionary dynamics through the differential inheritence of alleles over generations [@Pizzari2012; @Wright2013; @Wild2014; @Liao2015; @Reid2016]. For example, how frequently a focal female mates is a reproductive decision that can directly and indirectly affect her fitness, and potentially alter the distribution of genotypes and patterns of inheritence within a population [e.g., @Cornell2007; @Diaz2010; @Okada2015; @Roberts2015a; @Worthington2016]. Her mating frequency and expression of mate choice might be affected by the genetic quality or complementarity of available potential mates, and her choosiness might additionally be expressed after mating has taken place through post-copulatory mechanisms [@Parker1983; @Parker2006; @Kuijper2012]. Female choice might therefore be realised by evolution of multiple, potentially interdependent, behavioural and physiological phenotypes operating in tandem. However, no general modelling framework predicts the conditions under which female mating frequency, pre-copulatory choice, and post-copulatory choice evolve together, meaning that there are no formal hypotheses for predicting the relevance of these reproductive decisions across different biological systems.

Inbreeding, defined as breeding between relatives, directly affects inheritence of identical-by-descent alleles and the viability of inbred offspring, meaning that the decision to prefer, avoid, or tolerate (i.e., mate randomly with respect to relatedness) inbreeding is predicted to have profound effects on individual fitness [@Parker2006; @Kokko2006; @Szulkin2012; @Duthie2015a]. Viability of inbred offspring typically decreases with increasing relatedness between parents [termed 'inbreeding depression'; @Keller2002; @Charlesworth2009]. If inbreeding depression is sufficiently weak, selection for female inbreeding preference is predicted because inbreeding increases the reproductive success of a focal female's male relative, thereby increasing the number of identical-by-descent allele copies present in her offspring [@Parker1979; @Parker2006; @Kokko2006; @Duthie2015a]. However, if inbreeding depression is moderate to strong, selection for female inbreeding avoidance is predicted [Duthie and Reid *in press*; @Duthie]. Female inbreeding strategy can be enacted before or after copulation occurs [e.g., @Tuni2013; @Firman2015; @Daniel2015; @Fischer2015]. However, models of inbreeding do not consider pre-copulatory inbreeding strategy (hereafter 'inbreeding mating strategy') and post-copulatory inbreeding strategy (hereafter 'inbreeding fertilisation strategy') jointly, meaning that it is unclear when selection for either or both strategies is predicted in empirical systems.

Critically, evolution of inbreeding fertilisation strategy is precluded if females only mate with one male, meaning that evolution of polyandry (defined as female multiple mating within a single reproductive bout) is intrinsically tied to female inbreeding fertilisation strategy. Although polyandry is commonplace in wild populations, the regularity with which it is observed in the wild remains theoretically puzzling because engaging in polyandry does not necessarily increase the number of offspring that a female produces, meaning that the fitness benefits associated with polyandry are not always clear [@Parker2013; @Taylor2014]. It has been hypothesised that females can acheive a net fitness benefit if polyandry provides them with additional opportunities to adjust their inbreeding through copulations with additional males [e.g., @Tregenza2002; @Akcay2007; @Lehtonen2015]. @Duthie showed that selection for polyandry to facilitate inbreeding avoidance is sometimes predicted, but only under highly constrained conditions in which very few males are available during a female's initial mate choice but many males are available during a female's additional mate choice(s), or when polyandry is conditionally expressed based on the degree to which a focal female engaged in inbreeding given her initially chosen mate. However, @Duthie assumed that polyandrous females had no control over the assignment of paternity to their offspring. If after engaging in polyandry, females could bias paternity toward non-relatives, then selection for polyandry to avoid inbreeding might be more widely predicted [@Zeh1997]. However, no models quantify the degree to which expression of biased paternity resulting from evolving inbreeding fertilisation strategy can feedback to drive evolution of polyandry in populations of potentially inbreeding females and males with biologically realistic distributions of relatedness.

Evolution of inbreeding fertilisation strategy might increase selection for polyandry, but evolutionary feedbacks between inbreeding fertilisation strategy and polyandry might also be contingent upon evolution of inbreeding mating strategy, thereby leading to complex evolutionary dynamics between inbreeding strategies and polyandry. For example, if females can acheive their desired degree of inbreeding through mate choice alone, then subsequent polyandry and inbreeding fertilisation strategy might be superfluous for increasing female fitness. Alternatively, if females are often unsuccessful at realising their inbreeding mating strategy due to factors such as sexual conflict over mating and resulting male coercion [e.g., @Parker2006], limited kin recognition amongst potential mates [e.g., @Penn2010], or simply the inherent stochasticity associated with mate encounters or expression of mate choice based on inbreeding [e.g., @Kokko2006; @Duthie], then selection on inbreeding fertilisation strategy might occur to reinforce inbreeding mating strategy. In empirical systems where inbreeding depression is severe and inbreeding avoidance is therefore beneficial, expression of inbreeding mating strategy and inbreeding fertilisation strategy have been observed across diverse taxa [e.g., @Gasparini2011; @Tuni2013; @Firman2015; @Fischer2015; @Daniel2015]. However, no models predict *a priori* which of these inbreeding strategies should be observed when inbreeding depression is sufficiently severe to make inbreeding avoidance adaptive.

Evolution of polyandry, inbreeding mating strategy, and inbreeding fertilisation strategy will ultimately be determined by the, potentially interacting, fitness benefits and costs of each phenotype. Given severe inbreeding depression, phenotypes will increase a focal female's fitness if they cause her to avoid breeding with close relatives [@Parker1979; @Parker2006; @Kokko2006; @Duthie2015a]. However, evolution of phenotypes will also be directly affected by negative selection (i.e., costs) associated with the tendency to engage in polyandry, inbreeding mating strategy, and inbreeding fertilisation strategy. For example, females with a tendency to engage in polyandry might pay a cost associated with searching or making themselves available to mates, thereby increasing their own risk of predation [e.g., @Rowe1988; @Ronkainen1994; @Koga1998]. Females that are particularly choosy in mating interactions (e.g., especially reluctant to mate with relatives) might pay a cost of increased risk of mating failure, or harm caused by sexual conflict [@Rowe1994; @Kokko2013]. Females that are choosy after mating has occurred might pay an energetic cost associated with physiological or biochemical mechanisms needed to store sperm and successfully bias fertilisation [e.g., @Gasparini2011; @Tuni2013; @Fitzpatrick2014b]. Overall, to predict evolution of polyandry and inbreeding strategies, it is therefore necessary to model costs associated with polyandry, inbreeding mating strategy, and inbreeding fertilisation strategy.

Selection for inbreeding mating strategy versus inbreeding fertilisation strategy and associated polyandry might also depend evolutionary constraints, or on existing phenotypes in the population. For example, if phenotypes biasing fertilisation to avoid inbreeding are precluded by evolutionary constraints [@Bradshaw1991; @Futuyma2010], then inbreeding avoidance might only be possible before copulation occurs, perhaps strengthening selection for inbreeding mating strategy and weakening selection for polyandry. Alternatively, evolution of female inbreeding mating strategy might be precluded due to severely limited kin recognition or the inability to resist male coercion given sexual conflict [@Arnqvist2005a; @Parker2006; @Penn2010]. In such conditions, females might only be able to express choice through inbreeding fertilisation strategy, perhaps strengthening selection for inbreeding avoidance through polyandry and post-copulatory mechanisms. Selection for inbreeding mating or fertilisation strategy might also be weak or negligible if the strategy is redundant due to the presence of an already established adaptive inbreeding strategy (e.g., if females already successfully avoid inbreeding during mate encounters, then they might be unable to increase their fitness further by biasing fertilisation after mating has occurred). It is therefore necessary to consider evolution of inbreeding mating strategy and inbreeding fertilisation strategy when each evolves in a population where its complementary inbreeding strategy is evolutionarily constrained or already fixed as an adaptation.

@@

Studies of multiple species have found evidence for female pre-copulatory inbreeding avoidance [e.g., @Lihoreau2007; @Kuriwada2011; @Kingma2013; @Fischer2015], and for post-copulatory inbreeding avoidance [e.g., @Simmons2006; @Bretman2009]. Further, evidence of both pre-copulatory and post-copulatory inbreeding avoidance has been found for a handful of well-studied species, including Trinidadian guppies [*Poecilia reticulata*; @Gasparini2011; Daniel2015] and house mice [*Mus domesticus*; @Potts1991; @Firman2015], while @Liu2014 found evidence for pre-copulatory, but not post-copulatory, inbreeding avoidance in an experimental study of cabbage beetles (*Colaphellus bowringi*). These empirical studies demonstrate that evolution of all combinations of pre-copulatory and post-copulatory inbreeding avoidance are possible, but no general modelling attempts to predict *a priori* whether or not inbreeding avoidance will be enacted before or after copulation, or if both pre-copulatory and post-copulatory inbreeding avoidance will evolve. Here we have shown that evolution of inbreeding strategies and associated polyandry are predicted to be affected by direct costs, and constraints and adaptations of complementary inbreeding strategies.

>>>>

---
title: 'Evolution of pre-copulatory and post-copulatory inbreeding behaviour'
author: "A. Bradley Duthie et al."
date: "August 24, 2016"
output:
  pdf_document:
    fig_caption: yes
  html_document: default
  word_document:
    fig_caption: yes
    pandoc_args:
    - --csl
    - evolution.csl
header-includes:
- \usepackage{natbib}
- \usepackage{lineno}
- \linenumbers
- \bibliographystyle{amnatnat}
linestretch: 1
csl: evolution.csl
bibliography: ms_pre_post.bib
---

@@

<!--- Reproductive decisions expressed among interacting individuals directly affect individual fitnesses and the genotypes and viabilities of resulting offspring, and thereby profoundly affect evolutionary dynamics through the differential inheritence of alleles over generations [@Pizzari2012; @Wright2013; @Wild2014; @Liao2015; @Reid2016]. For example, how frequently a focal female mates is a reproductive decision that can directly and indirectly affect her fitness, and potentially alter the distribution of genotypes and patterns of inheritence within a population [e.g., @Cornell2007; @Diaz2010; @Okada2015; @Roberts2015a; @Worthington2016]. Her mating frequency and expression of mate choice might be affected by the genetic quality or complementarity of available potential mates, and her choosiness might additionally be expressed after mating has taken place through post-copulatory mechanisms [@Parker1983; @Parker2006; @Kuijper2012]. Female choice might therefore be realised by evolution of multiple, potentially interdependent, behavioural and physiological phenotypes operating in tandem. However, no general modelling framework predicts the conditions under which female mating frequency, pre-copulatory choice, and post-copulatory choice evolve together, meaning that there are no formal hypotheses for predicting the relevance of these reproductive decisions across different biological systems. --->

Polyandry, defined as female multiple mating within a single reproductive bout, is a reproductive strategy that simultaneously allows females to choose multiple mates and creates the opportunity for further cryptic choice enacted through post-copulatory mechanisms. One key choice for females is the decision to prefer, avoid, or tolerate breeding with relatives ['inbreeding'; @Szulkin2012]. Inbreeding often reduces the fitness of inbred offspring [@Charlesworth1999; @Keller2002; @Charlesworth2009], and polyandry has been widely hypothesised to evolve as an adaptation to allow females to reduce their inbreeding and therefore produce higher fitness offspring [@Zeh1997; @Jennions2000; @Tregenza2002; @Akcay2007]. In a recent model, @Duthie showed that selection for alleles underlying polyandry and pre-copulatory inbreeding avoidance occurred only under highly constrained conditions in which direct costs of polyandry were low, and when very few males were available during a female's initial mate choice but many males were available during a female's additional mate choice(s), or when polyandry was conditionally expressed based on the degree to which a focal female engaged in inbreeding given her initially chosen mate. In the absence of these constraints, expression of polyandry was expected to increase inbreeding with each additional mate because remaining males were more likely to be relatives once available non-relatives were chosen. However, @Duthie assumed that polyandrous females had no control over the assignment of paternity to their offspring. If after engaging in polyandry, females could bias paternity through post-copulatory mechanisms, then selection for polyandry to avoid inbreeding might be more widely predicted when inbreeding depression is severe [@Zeh1997].

<!--- I fear we're starting down a rabbit hole here. ---> 
Severe inbreeding depression might cause selection for alleles underlying active pre- and post-copulatory inbreeding avoidance and associated polyandry, but if evolution of one strategy renders its complement redundant for avoiding inbreeding, then selection might instead be sustained over generations for only one strategy. Empirical studies of multiple species have found evidence for female pre-copulatory inbreeding avoidance [e.g., @Lihoreau2007; @Kuriwada2011; @Kingma2013; @Fischer2015], and for post-copulatory inbreeding avoidance [e.g., @Simmons2006; @Bretman2009]. Further, evidence of both pre-copulatory and post-copulatory inbreeding avoidance has been found for a handful of well-studied species, including Trinidadian guppies [*Poecilia reticulata*; @Gasparini2011; Daniel2015] and house mice [*Mus domesticus*; @Potts1991; @Firman2015], while @Liu2014 found evidence for pre-copulatory, but not post-copulatory, inbreeding avoidance in an experimental study of cabbage beetles (*Colaphellus bowringi*). These empirical studies demonstrate that all combinations of adaptive pre-copulatory and post-copulatory inbreeding avoidance are possible, but no conceptual framework exists to predict *a priori* what combinations should be favoured by selection when both pre- and post-copulatory inbreeding avoidance, and polyandry, can evolve. Frequencies of alleles underlying active pre- and post-copulatory inbreeding avoidance and polyandry are likely to be affected by direct negative selection on phenotypes (i.e., costs). For example, females that are genetically predisposed to polyandry often pay a cost associated with searching or making themselves available to mates, thereby increasing their own risk of predation [e.g., @Rowe1988; @Ronkainen1994; @Koga1998]. Females that express pre-copulatory choice (e.g., reluctance to mate with relatives) might pay a cost of increased risk of mating failure, or harm caused by sexual conflict [@Rowe1994; @Kokko2013]. Females that express post-copulatory choice might pay an energetic cost associated with physiological or biochemical mechanisms needed to store sperm and successfully bias fertilisation [e.g., @Gasparini2011; @Tuni2013; @Fitzpatrick2014b]. If the relative costs affecting pre- and post-copulatory inbreeding avoidance differ, then alleles underlying the less costly inbreeding strategy might become fixed in the population over generations, while alleles underlying the more costly inbreeding strategy might become extinct if their effects are both costly and made redundant by the less costly strategy.

In addition to the relative importance of costs, evolution of pre- or post-copulatory inbreeding avoidance might be precluded if inbreeding avoidance is already acheived through a complementary strategy. Initial evolution of pre- and post-copulatory inbreeding avoidance strategies is unlikely to be simultaneous, so to predict evolutionary outcomes, it is necessary to consider evolution when one inbreeding avoidance strategy is initially rare, and the other strategy is common. For example, if pre-adaptation or rapid selection resulted in the fixation of alleles underlying adaptive pre-copulatory inbreeding avoidance, then alleles underlying post-copulatory inbreeding avoidance might be precluded from invading a population in which post-copulatory choice is not already expressed because the effect that such invading alleles had on decreasing total inbreeding avoidance would be weak or nelgigible.

<!--- It is therefore necessary to consider evolution of inbreeding mating strategy and inbreeding fertilisation strategy when each evolves in a population where its complementary inbreeding strategy is evolutionarily constrained or already fixed as an adaptation. --->

<!--- Overall, to predict evolution of polyandry and inbreeding strategies, it is therefore necessary to model costs associated with polyandry, inbreeding mating strategy, and inbreeding fertilisation strategy. --->

<!--- Inbreeding, defined as breeding between relatives, directly affects inheritence of identical-by-descent alleles and the viability of inbred offspring, meaning that the decision to prefer, avoid, or tolerate (i.e., mate randomly with respect to relatedness) inbreeding is predicted to have profound effects on individual fitness [@Parker2006; @Kokko2006; @Szulkin2012; @Duthie2015a]. Viability of inbred offspring typically decreases with increasing relatedness between parents [termed 'inbreeding depression'; @Keller2002; @Charlesworth2009]. If inbreeding depression is sufficiently weak, selection for female inbreeding preference is predicted because inbreeding increases the reproductive success of a focal female's male relative, thereby increasing the number of identical-by-descent allele copies present in her offspring [@Parker1979; @Parker2006; @Kokko2006; @Duthie2015a]. However, if inbreeding depression is moderate to strong, selection for female inbreeding avoidance is predicted [Duthie and Reid *in press*; @Duthie]. Female inbreeding strategy can be enacted before or after copulation occurs [e.g., @Tuni2013; @Firman2015; @Daniel2015; @Fischer2015]. However, models of inbreeding do not consider pre-copulatory inbreeding strategy (hereafter 'inbreeding mating strategy') and post-copulatory inbreeding strategy (hereafter 'inbreeding fertilisation strategy') jointly, meaning that it is unclear when selection for either or both strategies is predicted in empirical systems.  --->

<!--- Critically, evolution of inbreeding fertilisation strategy is precluded if females only mate with one male, meaning that evolution of polyandry (defined as female multiple mating within a single reproductive bout) is intrinsically tied to female inbreeding fertilisation strategy. Although polyandry is commonplace in wild populations, the regularity with which it is observed in the wild remains theoretically puzzling because engaging in polyandry does not necessarily increase the number of offspring that a female produces, meaning that the fitness benefits associated with polyandry are not always clear [@Parker2013; @Taylor2014]. It has been hypothesised that females can acheive a net fitness benefit if polyandry provides them with additional opportunities to adjust their inbreeding through copulations with additional males [e.g., @Tregenza2002; @Akcay2007; @Lehtonen2015].  However, no models quantify the degree to which expression of biased paternity resulting from evolving inbreeding fertilisation strategy can feedback to drive evolution of polyandry in populations of potentially inbreeding females and males with biologically realistic distributions of relatedness. --->

<!--- Evolution of inbreeding fertilisation strategy might increase selection for polyandry, but evolutionary feedbacks between inbreeding fertilisation strategy and polyandry might also be contingent upon evolution of inbreeding mating strategy, thereby leading to complex evolutionary dynamics between inbreeding strategies and polyandry. For example, if females can acheive their desired degree of inbreeding through mate choice alone, then subsequent polyandry and inbreeding fertilisation strategy might be superfluous for increasing female fitness. Alternatively, if females are often unsuccessful at realising their inbreeding mating strategy due to factors such as sexual conflict over mating and resulting male coercion [e.g., @Parker2006], limited kin recognition amongst potential mates [e.g., @Penn2010], or simply the inherent stochasticity associated with mate encounters or expression of mate choice based on inbreeding [e.g., @Kokko2006; @Duthie], then selection on inbreeding fertilisation strategy might occur to reinforce inbreeding mating strategy. In empirical systems where inbreeding depression is severe and inbreeding avoidance is therefore beneficial, expression of inbreeding mating strategy and inbreeding fertilisation strategy have been observed across diverse taxa [e.g., @Gasparini2011; @Tuni2013; @Firman2015; @Fischer2015; @Daniel2015]. However, no models predict *a priori* which of these inbreeding strategies should be observed when inbreeding depression is sufficiently severe to make inbreeding avoidance adaptive. --->

<!--- Evolution of polyandry, inbreeding mating strategy, and inbreeding fertilisation strategy will ultimately be determined by the, potentially interacting, fitness benefits and costs of each phenotype. Given severe inbreeding depression, phenotypes will increase a focal female's fitness if they cause her to avoid breeding with close relatives [@Parker1979; @Parker2006; @Kokko2006; @Duthie2015a]. However,  --->

<!--- Selection for inbreeding mating strategy versus inbreeding fertilisation strategy and associated polyandry might also depend evolutionary constraints, or on existing phenotypes in the population. For example, if phenotypes biasing fertilisation to avoid inbreeding are precluded by evolutionary constraints [@Bradshaw1991; @Futuyma2010], then inbreeding avoidance might only be possible before copulation occurs, perhaps strengthening selection for inbreeding mating strategy and weakening selection for polyandry. Alternatively, evolution of female inbreeding mating strategy might be precluded due to severely limited kin recognition or the inability to resist male coercion given sexual conflict [@Arnqvist2005a; @Parker2006; @Penn2010]. In such conditions, females might only be able to express choice through inbreeding fertilisation strategy, perhaps strengthening selection for inbreeding avoidance through polyandry and post-copulatory mechanisms. Selection for inbreeding mating or fertilisation strategy might also be weak or negligible if the strategy is redundant due to the presence of an already established adaptive inbreeding strategy (e.g., if females already successfully avoid inbreeding during mate encounters, then they might be unable to increase their fitness further by biasing fertilisation after mating has occurred). --->

@@

<!--- Studies of multiple species have found evidence for female pre-copulatory inbreeding avoidance [e.g., @Lihoreau2007; @Kuriwada2011; @Kingma2013; @Fischer2015], and for post-copulatory inbreeding avoidance [e.g., @Simmons2006; @Bretman2009]. Further, evidence of both pre-copulatory and post-copulatory inbreeding avoidance has been found for a handful of well-studied species, including Trinidadian guppies [*Poecilia reticulata*; @Gasparini2011; Daniel2015] and house mice [*Mus domesticus*; @Potts1991; @Firman2015], while @Liu2014 found evidence for pre-copulatory, but not post-copulatory, inbreeding avoidance in an experimental study of cabbage beetles (*Colaphellus bowringi*). These empirical studies demonstrate that evolution of all combinations of pre-copulatory and post-copulatory inbreeding avoidance are possible, but no general modelling attempts to predict *a priori* whether or not inbreeding avoidance will be enacted before or after copulation, or if both pre-copulatory and post-copulatory inbreeding avoidance will evolve. Here we have shown that evolution of inbreeding strategies and associated polyandry are predicted to be affected by direct costs, and constraints and adaptations of complementary inbreeding strategies. --->
